Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping

### 2024-06-24 paper/paper.tex
Updated to address reviewer comments for first round of review

Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters, e.g. from UAV-captured hyperspectral images (HSI). On-board computers installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}. These band ratios and polynomial combinations of bands have been used to successfully invert optically active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}. Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a wide range of parameters such as colored dissolved organic matter, chlorophyll A, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}. The calibration and evaluation of these data-driven models require a significant volume of coincident in situ data. This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}. However, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation. The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.

@@

A Freefly Alta-X autonomous quadcopter was used as the UAV platform in this study. This UAV was equipped with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager to capture HSI with $462$ wavelengths per pixel ranging from $391$ to $1011$ nm. This imager is in a pushbroom configuration so that HSI are captured one scan-line at a time resulting in data cubes consisting of $1000$ scan-lines with $1600$ pixels each. Additionally, the camera includes an embedded GPS/INS unit to enable georectification of collected HSI. An upward facing Ocean Optics UV-Vis NIR spectrometer with a cosine corrector was also included to provide measurements of the incident solar irradiance spectrum. The configuration of the UAV with the attached hyperspectral imager is shown in Fig~\ref{fig:drone}. Data collection and processing was controlled by an attached Intel NUC small-form-factor computer.

@@

A second GTM was trained on the combined dataset of HSI with pixels covering shore, water, and the rhodamine tracer dye release. The resulting \textcolor{red}{latent space distribution} is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the separation of distinct sources obtained by the GTM. \textcolor{red}{We note that a portion of the near-shore algae was observed to float at the surface of the water near the shore. Consequently, the position of the exemplar spectrum in Fig~\ref{fig:gtm-ndvi} used for algae occurs at a non-negative NDVI. The accumulation of records with negative NDVI values in close proximity (in the latent space) to the selected algae spectrum reflects the presence of algae within the water.}

@@

\textcolor{red}{Examining the latent space distribution learned by the GTM provides an unsupervised method to extract endmember spectra corresponding to unique sources within the imaging scene. To demonstrate this clearly, we consider the exemplar spectra visually identified in Fig~\ref{fig:sample-spectra}.} Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$. The responsibilities $R_{kn}$ therefore correspond to the contributions of the $kth$ node $\xi_k$ to the $n$th sample spectrum in the dataset. \textcolor{red}{Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}.} We note that these extracted endmembers are able to accurately capture the shape of each exemplar spectrum including thin reflectance peaks. The GTM has also managed to smoothly interpolate through noisy wavelengths as shown by the water and algae endmembers for $\lambda > 800$ nm.

@@

Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by identifying pixels with a NS3 value below a given threshold. In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond. \textcolor{red}{The distribution of NS3 values was multi-modal with a majority of pixels above $0.4$ belonging to dissimilar spectra. Consequently, a cutoff threshold of $0.4275$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$. Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights. \textcolor{red}{Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution.} In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$. A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$. The increase in NS3 values between flights reflects the dilution of dye as it diffused into the water.

@@

The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform a semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}. For example, in clear waters where variation in sediments and vegetation at the pond floor will contribute to observed spectral variability, the distribution learned by the GTM can be paired with in-situ data to classify the surface type.

@@

In this study we present the GTM as a useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers. Using data collected at a North Texas pond, we demonstrate how the latent space of the GTM can be used to visualize the distribution of observed reflectance spectra revealing the small-scale spatial variability of water composition. Spectral signatures extracted from GTM nodes are used to successfully map the abundance of algae near the shore and to track the evolution of a rhodamine tracer dye plume. These examples illustrate the power of combining unsupervised learning with UAV-based HSI collection for the characterization of water composition.

>>>>

Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI). On-board computers installed alongside hyperspectral imagers enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}. These band ratios and polynomial combinations of bands have been used to successfully invert optically active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}. Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a wide range of parameters such as colored dissolved organic matter, chlorophyll A, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}. The calibration and evaluation of these data-driven models requires a significant volume of coincident in situ data. This can be streamlined by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}. New proximal sensors for water quality parameters are continually being developed \cite{proximal-sensing-1, proximal-sensing-2}. However, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation. The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.

@@

A Freefly Alta-X autonomous quadcopter was used as the UAV platform in this study. This UAV was equipped with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager to capture HSI with $462$ wavelengths per pixel ranging from $391$ to $1011$ nm. This imager is in a pushbroom configuration so that HSI are captured one scan-line at a time resulting in data cubes consisting of $1000$ scan-lines with $1600$ pixels each. Additionally, the camera includes an embedded GPS/INS unit to enable georectification of collected HSI. An upward facing Ocean Optics UV-Vis NIR spectrometer with a cosine corrector was also included to provide measurements of the incident solar irradiance spectrum. The configuration of the UAV with the attached hyperspectral imager is shown in Fig~\ref{fig:drone}. Data collection and processing was controlled by an attached Intel NUC small-form-factor computer. A detailed description of the UAV platform can be found in \cite{robot-team-1}.

@@

A second GTM was trained on the combined dataset of HSI with pixels covering shore, water, and the rhodamine tracer dye release. The resulting latent space distribution is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the separation of distinct sources obtained by the GTM. We note that a portion of the near-shore algae was observed to float at the surface of the water near the shore. Consequently, the position of the exemplar spectrum in Fig~\ref{fig:gtm-ndvi} used for algae occurs at a non-negative NDVI. The accumulation of records with negative NDVI values in close proximity (in the latent space) to the selected algae spectrum reflects the presence of algae within the water.

@@

Examining the latent space distribution learned by the GTM provides an unsupervised method to extract endmember spectra corresponding to unique sources within the imaging scene. To demonstrate this clearly, we consider the exemplar spectra visually identified in Fig~\ref{fig:sample-spectra}. Using the trained GTM we can compute spectral signatures for each node in the latent space via the non-linear mapping $\psi$. The responsibilities $R_{kn}$ therefore correspond to the contributions of the $kth$ node $\xi_k$ to the $n$th sample spectrum in the dataset. Endmembers are identified by nodes with non-zero responsibility values for each of the selected exemplar spectra and are plotted in Fig~\ref{fig:spectral-signatures}. We note that these extracted endmembers are able to accurately capture the shape of each exemplar spectrum including thin reflectance peaks. The GTM has also managed to smoothly interpolate through noisy wavelengths as shown by the water and algae endmembers for $\lambda > 800$ nm.

@@

Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by identifying pixels with a NS3 value below a given threshold. In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond. The distribution of NS3 values was multi-modal with a majority of pixels above $0.4$ belonging to dissimilar spectra. Consequently, a cutoff threshold of $0.4275$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution. The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$. Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights. Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution. In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$. A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$. The increase in NS3 values between flights reflects the dilution of dye as it diffused into the water.

@@

The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumptions of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform a semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}. For example, in clear waters where variation in sediments and vegetation at the pond floor will contribute to observed spectral variability, the distribution learned by the GTM can be paired with in-situ data to classify the surface type.

@@

In this study we present the GTM as a useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers. Using data collected at a North Texas pond, we demonstrate how the latent space of the GTM can be used to visualize the distribution of observed reflectance spectra revealing the small-scale spatial variability of water composition. Spectral signatures extracted from GTM nodes are used to successfully map the abundance of algae near the shore and to track the evolution of a rhodamine tracer dye plume. These examples illustrate the power of combining unsupervised learning with UAV-based hyperspectral imaging for the characterization of water composition. Future work will further develop the GTM as a tool to guide in-situ data collection and enable contaminant localization for real-time applications.
